Extended Data

### 2021-10-18 doc/manuscript/Dow_manuscript_Nature.Rmd
methods

Both leaf phenology and stem growth were sensitive to spring temperatures (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-6), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan. 1 and $DOY_{25}$ for which the correlation between $T_{max}$ and $DOY_{25}$ was strongest (henceforth, critical temperature window; Fig. `r schematic_summary`). 
The strongest relationships, including but not limited to the critical temperature window (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$ (`r EDF_climwin`).

@@

The critical temperature window (Fig. `r schematic_summary`), defined as the period over which $DOY_{25}$ was most strongly affected by early spring $T_{max}$, was determined using the R package *climwin*[@vandepol_identifying_2016]. 
This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and difference in Akaike information criterion corrected for small sample size relative to the null model, $\Delta AICc$, of each window. 
Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group (by xylem porosity) and site (`r  EDT_mean_growth_parameters`), selecting the critical temperature window as that with the lowest $\Delta AICc$.
Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable, as opposed to $T_{min}$.
We focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters, because we were interested in the influence of temperature on the initiation of woody growth,

@@

(*This section needs work/ more detail.*)

Correlation between these variables and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.

Mixed effects models were used to test the response of growth phenology variables ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_max}$, $g_{max}$, and $\Delta DBH$) to fixed effects of xylem porosity and mean temperature ($T_{max}$, $T_{min}$) during the critical temperature window selected by *climwin*, along with random effects of species and tree. 
We ran separate models for each species group at each site, and for the response of all growth phenology variables to $T_{max}$ and $T_{min}$. 
Models were run using the LME4 package in R [@bates_fitting_2015].

This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language @rstanarm_2020. In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 
<!--
TODO added by Bert: Whereever we specify model details, include details about prior from here:
https://cran.r-project.org/web/packages/rstanarm/vignettes/priors.html#default-weakly-informative-prior-distributions
-->

>>>>

Both leaf phenology and stem growth were sensitive to spring temperatures (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-5), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan. 1 and $DOY_{25}$ for which the correlation between $T_{max}$ and $DOY_{25}$ was strongest (henceforth, critical temperature window; Fig. `r schematic_summary`). 
The strongest relationships, including but not limited to the critical temperature window (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$ (`r EDF_climwin`).

@@

The critical temperature window (Fig. `r schematic_summary`), defined as the period over which $DOY_{25}$ was most strongly affected by early spring $T_{max}$, was determined using the R package *climwin*[@vandepol_identifying_2016]. 
This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and difference in Akaike information criterion corrected for small sample size relative to the null model, $\Delta AICc$, of each window. 
Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group (by xylem porosity) and site (`r  EDT_mean_growth_parameters`), selecting the critical temperature window as that with the lowest $\Delta AICc$.
Because $T_{max}$ proved to have a generally stronger influence over $DOY_{25}$ and other growth parameters, we focused on this variable, as opposed to $T_{min}$.
We focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters, because we were interested in the influence of temperature on the initiation of woody growth.

Correlation between growth parameters ($DOY_{25}$,$DOY_{50}$,$DOY_{75}$,$DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig. `r schematic_summary`) and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.
Analyses were run for both $T_{max}$ and $T_{min}$, with qualitatively similar results, but we present only results for $T_{max}$, which had overall stronger correlation with growth parameters. 
Mixed effects models were used to test the response of growth phenology variables to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the critical temperature window selected by *climwin*, along with random effects of species and tree. 
We ran separate models for each species group at each site, and for the response of all growth phenology variables to $T_{max}$ (or $T_{min}$). 
Models were run using the LME4 package in R [@bates_fitting_2015]. 
This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language @rstanarm_2020. In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 
<!--
TODO added by Bert: Whereever we specify model details, include details about prior from here:
https://cran.r-project.org/web/packages/rstanarm/vignettes/priors.html#default-weakly-informative-prior-distributions
-->

@@
